---
title: "prevR Vignette (Short)"
author: "Stephanie J. Lax"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rprev Vignette Short}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This document demonstrates how predictions of prevalence are calculated from a synthetic patient dataset. 
```{r backgroundsetup, echo = FALSE, message = FALSE}
library(devtools)
devtools::load_all()
```

## Basic Experimental Setup

```{r setup, message = FALSE, warning = FALSE}
library(dplyr)
library(abind)
library(rms)
library(survival)
library(knitr)
library(tidyr)
library(ggplot2)
library(dplyr)
load("H:/Repos/prevR/data/prevsim.rda")
registry_data <- prevsim
N_years <- 10
cure <- 3

summary(registry_data)
```

```{r basicsurvival, fig.height=4, fig.width=7}
survf <- survfit(Surv(time, status) ~ sex, data=registry_data)
plot(survf, lwd=2, col = c(1:2), xlab="survival (days)", ylab="probability")
legend(3000, 1, c("Males", "Females"), lty = 1, col=c(1:2))
```

*prevsim* is a synthesised dataset to resemble registry data, such as that found in HMRN. Incident cases are recorded from `r min(registry_data$entrydate)` to `r max(registry_data$entrydate)`, and events occur between `r min(registry_data$eventdate)` and `r max(registry_data$eventdate)`. `cure` represents the survival model used in the predictions, i.e. 3 means a patient can be considered cured 3 years after diagnosis, and therefore from this point onwards their survival is modelled using population data rather than patient data. `N_years` corresponds to the number of preceding years for which prevalence is predicted. 

## Elementary Calculations

The following code calculates absolute incidence of disease for each selected year of the registry:

```{r totalincidence, error = TRUE}
raw_incidence <- incidence(registry_data$entrydate, start="2004-01-30", 9)
raw_incidence
```

It can be concluded that the disease has around 100 new cases each year. To assess consistency of the incidence data with a homogeneous Poisson process, the `sim_check` and `smoothed_incidence` functions can be used:

```{r}
sim_check(raw_incidence)
```

```{r, fig.width = 7, fig.height = 7, include = TRUE, results="hide"}
par(mfrow=c(2,2))
smoothed_incidence(registry_data$entrydate, start_date="2004-01-30", num_years=9)
```

The age distribution of incident cases can also be inspected:

```{r incidenceage, fig.width = 4, fig.height = 4, error = TRUE}
#par(mfrow=c(1,1))
#incidence_age_distribution(load_data(registry_data), registry_years, registry_start_year, 
#                           registry_end_year)
```

To calculate the mean incidence rate per 100,000 within the study population `meanIR` is used, which requires an estimate of the size of the population at risk:
  
```{r incidencerate, error = TRUE}
meanIR(registry_data$entrydate, population_size = 3500000,
                          start = "2004-01-30", num_years = 9)
```

Prevalence can be counted from the data. There is a discrepency between the last recorded incident case and the last date of follow-up owing to a systematic difference in incidence and survival data collection. All the survival data for the available incident cases is used for survival modelling, however the prevalence estimate works back by year from an "index date", which is the last date of the last year of registry data, as the number of incident cases must be complete for simulating prevalence. Therefore, `r max(registry_data$entrydate)` is the index date for this synthetic dataset. In `counted_prevalence()`, by default, all cases alive after the index date are censored. The user is permitted to specify their desired index date if require, but `counted_prevalence()` will no longer be comparable to `prevalence()` and the user will be warned about this if they try to do so.

```{r survival, error = TRUE}
counted_prevalence(registry_data$entrydate, 
                     registry_data$eventdate, 
                     registry_data$status, 
                     start="2004-01-30", 9)
```

## Survival Modelling

Firstly, we can inspect the survival data for consistency between years of the registry and with a Cox Proportional Hazards model using `survival_modelling_diagnostics`:

```{r survivaldiag, fig.width = 7, fig.height = 7, include = TRUE, results="hide"}
#par(mfrow=c(2,2))
#survival_modelling_diagnostics(load_data(registry_data), ages = c(55, 65, 75, 85, 100), 
#                               registry_years, registry_start_year, registry_end_year)
```

```{r survivaldiag2, echo = FALSE, fig.show="hide"}
#output <- survival_modelling_diagnostics(load_data(registry_data), 
#                                         ages = c(55, 65, 75, 85, 100), registry_years, 
#                                         registry_start_year, registry_end_year)
#output[[5]]
```

### Check Functional Form of Age:

```{r ageform2, fig.width = 7, fig.height = 4, error = TRUE}
#par(mfrow=c(1,2))
#functional_form_age(load_data(registry_data), registry_years, registry_start_year, 
#                    registry_end_year)
```

Next, general population period survival data for the UK is loaded and yearly rates are translated into daily rates by linear interpolation using `population_survival_rate()`.

```{r popsurv, fig.width = 6, fig.height = 4, error = TRUE}
load("../data/UKmortality.rda")

daily_survival_males <- population_survival_rate(rate ~ age, subset(UKmortality, sex == 0))
daily_survival_females <- population_survival_rate(rate ~ age, subset(UKmortality, sex == 1))

plot(daily_survival_males, type="l", col="blue", xlab="days", ylab="survival")
lines(daily_survival_females, col="red")
legend("topright", legend = c("Males", "Females"),
               bty = "n", lty = 1, col = c(4,2))
```

We model survival in our patient cohort using a Weibull distribution. This is calculated within the prevalence function, however, it is important to look at `summary(wb)` for diagnostic information and to assess the model fit. 

```{r weibull, error = TRUE}
registry_data_r <- registry_data[registry_data$entrydate >= "2004-01-30", ]
wb <- survreg(Surv(time, status) ~ age + sex, data=registry_data_r)
summary(wb)
```

## Prevalence Calculations

Prevalence can now be calculated:

```{r prevalencetotal, error = TRUE}
prevalence_total <- prevalence(Surv(time, status) ~ sex(sex) + age(age) + entry(entrydate),
                               registry_data, N_years = N_years, cure_time = cure*365, 
                               start='2004-01-30', num_years=9)

by_year_total <- prevalence_total[[1]]
by_year_total
```

`by_year_total` is a vector of raw prevalence per year of the registry. Whether the model is predicting good numbers for the recent years for which we have actual prevalence can be assessed using a chi squared test:

```{r test, error = TRUE}
prev_chisq(registry_data$entrydate, 
           registry_data$eventdate, 
           registry_data$status, 
           start="2004-01-30", 9,
           by_year = by_year_total)
```

Up until now, the simulated values of prevalent cases for the first years before the index date have been used. Here we replace these with the actual observed values:

```{r hist1, fig.width = 6, fig.height = 4, error = TRUE}
observed <- counted_prevalence(entry, events, status, start, num_years)
predicted <- rev(by_year[1:num_years])
by_year_total[1:num_years] <- rev(observed)
```

The age distribution of simulated prevalent cases can be viewed as a histogram and compared with counted prevalent cases:

```{r, fig.height = 4, fig.width = 7}
registry_data_r$status2 <- ifelse(registry_data_r$eventdate > "2013-01-30", 
                                  0, registry_data_r$status)
observed_post_age_dist_total <- 
  registry_data_r$age[registry_data_r$status2 == 0]

post_age_dist_total <- prevalence_total[[2]]

hist(observed_post_age_dist_total, 
     col=rgb(0.1, 0.1, 0.1, 0.5), xlim=c(0,100), ylim=c(0,0.045), 
     main = "", xlab = "age", prob = TRUE)
hist(post_age_dist_total, 
     col=rgb(0.8, 0.8, 0.8, 0.5), prob = TRUE, add=T)
legend("topleft", legend = c("Observed", "Predicted"), bty = "n", lty = 1, 
       col = c(rgb(0.1, 0.1, 0.1, 0.5),
               rgb(0.8, 0.8, 0.8, 0.5)))
```

The age distribution of prevalent cases in the predicted years only can also be retrieved using `ageoutput`:

```{r output2, error = TRUE}
#age_props <- prevalence_by_age(dist = post_age_dist_total, registry_end_year, N_years)
#age_props
```

As a way to estimate uncertainty of our survival model, i.e. sampling variation, the prevalence function calculates bootstrapped survival coefficients. `by_year_samples_total` contains the raw prevalence data for each bootstrapped sample:

```{r bysample, error = TRUE}
#by_year_samples_total <- prevalence_total[[3]]
#by_sample_total <- apply(by_year_samples_total, 2, sum)
```

`n_year_estimates` is a function to calculate the estimated number of *n-year* prevalent cases for any *n*. It returns the absolute numerical estimate as well as an estimate of the prevalence proportion and 95% confidence intervals for the prevalence proportion. 

```{r nyrest3, error = TRUE} 
#total3yr <- n_year_estimates(N_years = 3, registry_start_year = registry_start_year, 
#                             registry_end_year = registry_end_year, 
#                             the_samples = by_year_samples_total, by_year = by_year_total,
#                             population_size = 3500000)
#total3yr
```
