---
title: "Diagnostics for rprev"
author: "S J Lax and S E Lacy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{rprev User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Introduction

The [user guide](user_guide.html) highlighted the basic use of the prevalence simulation methodology, highlighting that the prevalence process is a result of two distinct processes: incidence and survival. The package allows for custom specification of both of these systems, and provides default implementations that do not rely on strict assumptions about the data and work well in many situations. This vignette acts as a guide to assessing the modelling assumptions made when using these default models, illustrating how the diagnostics provided with the package are used to check the validity of the assumptions that are made for simulation, and checking the agreement between observed and simulated prevalence estimates.

First, we will inspect the consistency of incidence data between years of the registry and determine if it can be appropriately modelled as an homogeneous Poisson process. Second, we will look at the consistency of the survival data between years of the registry and check the adequacy of a parametric model of survival. We emphasise that the user should check that their registry data does meet the required assumptions and if not, should understand that estimates of prevalence made using simulation based on that data may not be correct.

```{r setup, message = FALSE, warning = FALSE, echo=F}
library(rprev)
library(survival)
data(prevsim)
```

# Incidence

The simulation of incident cases in the years for which registry data is not available currently assumes that the incidence process is homogeneous Poisson. The package includes several functions for inspecting the incidence data, as well as diagnosing the validity of this assumption.

The primary function is `incidence`, which calculates several summary statistics of the disease incidence. It requires a vector of entry dates into the registry, along with the size of the population under consideration. The registry under interest can be subset to only count incident cases from specific years of interest, useful if there are doubts about the accuracy of the data, for example during the first year of operation. This function also fits a spline smooth to the incidence rate data.

```{r}
inc <- incidence(prevsim$entrydate, population_size=3.2e6, start='2005-09-01', num_reg_years=5)
```

This function returns an S3 object of class `incidence` with the following attributes.

```{r}
inc
names(inc)
```

The `summary.incidence` method displays the most useful information about the incidence process of the disease of interest.

```{r}
summary(inc)
```

The known yearly incidence rates are displayed starting at `start` date for `num_reg_years`, starting with the first year. For the simulated data set `prevsim`, there were 96 incident cases between 2005-09-01 and 2006-09-01. It can be concluded that the disease has around 100 new cases each year.

Below that is a summary of the cumulative diagnosis times (in days), useful as a quick check of the distribution of incident cases. The p-values of a simulation test indicating whether the yearly incidence estimates are under or over dispersed relative to a homogeneous Poisson process are also displayed. Inspection of the smoothed incidence function should reveal whether the problem is one of non-homogeneity (which may lead to inaccurate prevalence estimates) or of failure of the Poisson assumption (which may lead to inaccurate estimates of confidence intervals). 

Another useful incidence statistic is the mean incidence rate per 100,000 within the study population, which is obtained from the `mean` attribute. Confidence intervals are provided at the specified level (default is 95%):
  
```{r incidencerate, error = TRUE}
inc$mean
```

Alongside the p-values from the over/under dispersion test, `rprev` provides visual tools to assess the consistency of the incidence data with an homogeneous Poisson process.

`plot.incidence` displays a plot of the actual incidence rate as recorded in the registry, with the smooth overlaid. If incidence is an homogeneous Poisson process, both the smooth (green) and incidence process (red) should remain within the 95% confidence interval (dashed blue) and be evenly distributed about the mean (blue line). 

```{r, fig.width = 7, fig.height = 4}
plot(inc)
```

Another function used to describe the incident cases is `incidence_age_distribution`, which simply plots the distribution of incident cases by age.

```{r incidenceage, fig.width = 7, fig.height = 4, error = TRUE}
incidence_age_distribution(prevsim$age)
```


# Survival modelling

The second component to the prevalence estimation is the survival modelling process. In the default implementation, survival is modelled using a parametric fit. It is recommended that the user inspects the survival data for consistency with the assumptions of this model and for homogeneity. In this example, we'll use the `prevsim` data set and test the assumptions of a Weibull model acting on the sex and age covariates, since these are common demographic factors that are likely to be used in prevalence estimation.

First, it is always useful to plot the Kaplan-Meier estimator of the data, both as a whole and stratified by age to visually inspect for any inconsistencies:

```{r survivaldiag, fig.width = 7, fig.height = 4}
km <- survfit(Surv(time, status) ~ 1, data=prevsim)
plot(km, lwd=2, col="blue", main="Overall Survival", xlab="Days", 
     ylab="Survival probability")
```

```{r survivaldiag2, fig.width = 7, fig.height = 4}
ages = c(55, 65, 75, 85, 100)
km2 <- survfit(Surv(time, status) ~ cut(age, breaks=ages), data=prevsim)
plot(km2, lwd=2, col=1:length(ages), main="Survival stratified by age", xlab="Days", 
     ylab="Survival probability")
legend("topright", legend=substring(names(km2$strata), 25, 32), lty = 1, 
       col=1:length(ages))
```

It is also a useful diagnostic aid to plot the survival curve for each year of the registry to determine whether there is any inhomogeneity:

```{r survivaldiag4, fig.width = 7, fig.height = 4, results='hide'}
plot(km, lwd=2, col="blue", mark.time=F, conf.int=T, xlab="Days", 
     ylab="Survival probability")
num_reg_years <- 9
registry_years <- sapply(0:9, function(x) as.Date(paste0(2004+x, "-01-30")))
sapply(seq(num_reg_years),
       function(i) lines(survfit(Surv(time, status) ~ 1, 
                                 data=prevsim[prevsim$entrydate >= 
                                                          registry_years[i] & 
                                                          prevsim$entrydate < 
                                                          registry_years[i + 1], ]), 
                         mark.time = F, conf.int = F))
```

The effect of age on hazard can be visualised to determine if there are any non-proportional effects, by inspecting Schoenfeld residuals from a Cox model. This is easily done using the `cox.zph` function from the `survival` package. 

```{r survivaldiag3, fig.width = 7, fig.height = 4}
cx <- coxph(Surv(time, status) ~ age, data=prevsim)
cxp <- survfit(cx, 
               newdata=data.frame(age=sapply(seq(length(ages) - 1), 
                                             function(i) mean(c(ages[i], ages[i + 1]))))) 
plot(cox.zph(cx))
lines(cxp, lwd=2, col=1:length(ages), lty=2, mark.time=F)

```

An overall test of the proportional hazards assumption may be made.

```{r}
cox.zph(cx)
```

## Functional form of age

`functional_form_age()` is provided so that the user can further investigate if modelling the effect of age as linear is appropriate. This function fits a Cox model to the data with a restricted cubic spline with the specified degrees of freedom. The model is returned by the function as an `rms::cph object`, allowing the user to inspect the fit. The function can also plot relative hazard as a function of age with the `plot_fit` argument.

```{r ageform2, fig.width = 7, fig.height = 4, error = TRUE}
functional_form_age(Surv(time, status) ~ age, prevsim, df=4, plot_fit=T)
```

# Prevalence estimates

As a reminder, prevalence is estimated using incidence and survival data from *n* years. However, registry data (and thus known incidence and survival) data may only be known for *r* years, where *r* <= *n*. If *r* < *n*, the remaining *n*-*r* years of incidence and survival are simulated using Monte Carlo techniques. If the incidence and survival models are well specified then the prevalence estimates should be reliable, however, it is still useful to check the fit before drawing any conclusions from the results. 

```{r prevalencetotal, error = TRUE}
prevalence_total <- prevalence(index='2013-01-30', 
                               num_years_to_estimate=c(3, 5, 10, 20), 
                               data=prevsim, 
                               inc_formula = entrydate ~ sex,
                               surv_formula = Surv(time, status) ~ age + sex, 
                               dist='weibull', 
                               population_size = 1e6,
                               death_column = 'eventdate')
```

## Diagnosing incidence models

We do not provide any functions for diagnosing the incidence models fitted to the bootstrap data, however, all the objects are available in the `inc_models` attribute of the returned `prevalence` object and can be used to check for any errors in fitting.

## Diagnosing survival models

To inspect the distribution of the bootstrapped survival models [see @crouch2014determining for details], a `survfit.prev` object can be constructed using the usual `survfit` call, accepting a data frame of new data (with identical column names to those found in the original data set). In the example below, survival probability is estimated for a 60 year old male:

```{r}
prevsurv <- survfit(prevalence_total, 
                    newdata=data.frame(age=60, sex=factor('M', levels=levels(prevsim$sex))))
prevsurv
```

The `summary.survfit.prev` method provides *N*-year survival probabilities, with *N* specified as an argument vector:

```{r}
summary(prevsurv, years=c(1, 3, 5, 10))
```

Plotting the `survfit.prev` object displays the survival curve of a Weibull model using the original data set (orange), along with 95% confidence intervals derived using the bootstrapped models shaded in light grey, while outlier curves are individually plotted. Outliers are defined as those survival curves where at least `pct_show` proportion of their point predictions lie outside the 95% confidence interval. This plot is useful to assess the variability of the survival models. Further manual inspection can be carried out by looking at the objects themselves, saved in the `surv_models` attribute of the `prevalence` object.

```{r, fig.width=7, fig.height=4}
plot(prevsurv, pct_show=0.90)
```

## Comparison between simulated and counted prevalence

As a test of whether the model is predicting reasonable values of prevalence, we can use the fact that we can directly measure the discrepancy between the predicted and actual prevalence for the available registry years. This difference can be formally tested with an exact Poisson test of the counted prevalence from both the simulated estimate and the known registry value; the resulting p-value resulting is returned as an attribute of a `prevalence` object, called `pval`.

```{r}
prevalence_total$pval
```

For this model, there is no evidence to reject the null hypothesis.

This can also be calculated manually with the `test_prevalence_fit` function.

```{r test, error = TRUE}
test_prevalence_fit(prevalence_total)
```

## Simulated population

The `simulated` attribute of the `prevalence` object holds a `data.table` with the incident population from every simulation, along with derived fields indicating whether they contributed to prevalence at the index date of any years of interest. It can be used as an overall check of the appropriateness of the generated incidence population, as well as identifying any discrepencies in the survival modelling.

```{r}
knitr::kable(head(prevalence_total$simulated))
```


# References
